Fail Faster

### 2025-03-22 paper/paper.tex
rg
bind explanation

The basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
The constant generator \texttt{return} runs at compile time. Given \texttt{cx : 'a code}, the code for an \texttt{'a}, it
returns the generator that runs \texttt{cx}. Similarly,
\texttt{bind g k} sequences generators as usual by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one
% \footnote{Readers familiar with OCaml may notice that \texttt{return : 'a code -> 'a Gen.t} and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t} do not have the correct types for a monad instance, preventing us from using \texttt{let\%bind} notation. We rectify this issue in Section~\ref{subsection:codecps} by importing some clever ideas from the staging literature.}.
\jwc{Fix explanation here: bcp says ``this explanation doesn't quite explain what's going on. after all, in a staged program the contnuatino k doesn't know which values is going to be passed to it. the difference in the staged setting is that \texttt{k} is not allowed to ask any questions about its argument, it has to treat it totally parametrically.}
Operationally, bind takes code for the size and seed and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g} and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size\_c random\_c}
and \texttt{k .<a>. size\_c random\_c} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
from Section~\ref{subsection:msp}).
\iflater
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}
\fi

@@

\subsection{Recursive Generators}
\jwc{Generating recursive datatypes requires recursive generators!}

Different generator DSLs handle defining recursive generators differently. Some allow recursive generators
to be defined as recursive functions, while others
(including both \bq and ScalaCheck) expose a fixpoint combinator to construct recursive generators.
Given a step function that takes a ``handle'' to sample from a recursive generator call, it
ties the knot and builds a recursive generator.

>>>>

The basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
The constant generator \texttt{return} runs at compile time. Given \texttt{cx : 'a code}, the code for an \texttt{'a}, it
returns the generator that runs \texttt{cx}. Similarly,
\texttt{bind g k} sequences generators as usual by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
we know that at run time the code \texttt{g} generates will produce \emph{some} \texttt{'a}, but at compile time we cannot inspect the value.
Operationally, bind takes code for the size and seed and returns code that (1) let-binds a variable \texttt{a} to spliced-in code of \texttt{g} and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size\_c random\_c}
and \texttt{k .<a>. size\_c random\_c} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
from Section~\ref{subsection:msp}).
\iflater
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}
\fi

@@

\subsection{Recursive Generators}
Generating recursive datatypes requires recursive generators. Different generator DSLs handle defining recursive generators differently. Some allow recursive generators
to be defined as recursive functions, while others
(including both \bq and ScalaCheck) expose a fixpoint combinator to construct recursive generators.
Given a step function that takes a ``handle'' to sample from a recursive generator call, it
ties the knot and builds a recursive generator.
